---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

# spatialphy

<!-- badges: start -->

<!-- badges: end -->

`spatialphy` is an R package for spatial phylogentic analysis. The field of spatial phylogenetics is focused on accounting for evolutionary relationships among taxa when describing biodiversity patterns, which has a number of advantages over species-based accounting. This library currently includes functions for computing:

-   diversity measures like phylogenetic diversity
-   turnover measures like phylogenetic betaendemism
-   phylogenetic regionalization
-   phylogenetic conservation prioritization

The package is still actively under development; there may be bugs, and the API is subject to change.


## Installation

This package is not yet on CRAN. You can install the development version of spatialphy like so:

``` r
devtools::install_github("matthewkling/spatialphy")
```

## Examples

Spatial phylogenetic analyses require two essential ingredients: data on the geographic distributions of a set of organisms, and a phylogeny representing their evolutionary relationships. All workflows in the spatialphy package will involve creating a dataset of class `'spatialphy'` containing these necessary components. 

A user would normally define this object with their own data using the `sphy()` function, but we can also simulate a demonstraiton dataset using `simulate_sphy()`, which we'll do here. The dataset is a list containing:

- a phylogeny
- a matrix of occrrence data for evey clade at every level of the tree (this is constructed internally from user-supplied data about the occurrences of the terminal taxa)
- some spatial metadata, in the form of a raster layer

```{r init, message=FALSE, warning=FALSE}
library(spatialphy)
library(tmap); library(magrittr) # for visualization
set.seed(123)
sp <- simulate_sphy(n_tips = 50)
```


#### Alpha diversity

We can then pass this data object to the other functions in the library. The `diversity()` function calculates a number of alpha diversity measures, including phylogenetic diversity (PD), phylogenetic endemism (PE), and several others. 

```{r alpha, message=FALSE, warning=FALSE}
div <- diversity(sp)
names(div)
tm_shape(div$PD) + tm_raster()
```


#### Beta diversity (turnover)

We can also look at patterns in the phylogenetic similarity among locations. The first step here is to compute the community phylogenetic distance between every pair of sites; we'll use the `phylo_dist()` function to add a pariwise distance matrix as an additional component in our spatialphy dataset.

```{r beta, message=FALSE, warning=FALSE}
sp <- phylo_dist(sp, add = T)
```

Having done this, we can then assess spatial turnover patterns in a couple ways. We can qualitatively visualize them by converting the distance matrix to a set of colors representing how similar two sites are to each other, using the `phylo_rgb()` function:

```{r rgb, message=FALSE, warning=FALSE}
sp %>%
      phylo_rgb() %>%
      tm_shape() +
      tm_rgb(max.value = 1) +
      tm_layout(title = "Phylogenetic ordination")
```

We can also perform a more formal cluster analysis that splits the landscape into a set of evolutionary bioregions, using the `phylo_regions()` function. We'll ask for eight regions:

```{r regions, message=FALSE, warning=FALSE}
sp %>%
      phylo_regions(k = 8) %>%
      tm_shape() +
      tm_raster(palette = "cat") +
      tm_layout(legend.outside = T,
                title = "Phylogenetic regions")
```


#### Conservation prioritization

Finally, this package can also perform a basic spatial conservation prioritization, identifying priority locations for the creation of new protected areas in order to maximize biodiversity protection. This is done using the `prioritize()` function, which uses a greedy forward stepwise algorithm to produce a nested ranking of conservation priorities.

In addition to our spatial phylogenetic dataset, we could optionally supply a raster layer indicating the locations and effectiveness of existing protected areas. We could also modify parameters that specify the effective level of protection provided by new reserves, as well as the shape of the "benefit" function that determines the relative importance of protecting the initial populations of every taxon versus more populations of more phylogenetcally distinct taxa. But here we'll stick with the defaults. In the map below, the highest-priority sites are in yellow.

```{r priorities, message=FALSE, warning=FALSE}
sp %>%
      prioritize() %>%
      tm_shape() + 
      tm_raster(palette = c("yellow", "red", "blue", "black"),
                colorNA = "gray80", n = 20,
                title = "conservation priorities") + 
      tm_layout(legend.outside = T)
```

